\input{Preamble.tex}

\begin{document}
%\input{acronyms}
% Title Page
%\comment{
\date{\today}
\makeatletter
\begin{titlepage}
\centering
\vfill
Thesis submitted for the degree of Doctor of Philosophy (PhD)
\par
\vspace*{1in}
\begin{Large}\bfseries
The Economics of Type 2 Diabetes in Middle-Income Countries\par
\end{Large}
\vspace{1in}
\begin{large}\bfseries
Till Seuring\par
\end{large}
\par
\vspace{0.2in}
Norwich Medical School
\par
University of East Anglia, UK
\par
\vspace{1.0in}
September 2016
\par
\vspace{2.0in}
This copy of the thesis has been supplied on condition that anyone who consults it is
understood to recognise that its copyright rests with the author and that use of any
information derived there from must be in accordance with current UK Copyright Law.
In addition, any quotation or extract must include full attribution.
\end{titlepage}
\makeatother

\cleardoublepage
\addcontentsline{toc}{chapter}{\nameref{abstract}}
\chapter*{\label{abstract}Abstract}
\input{abstract}

\tableofcontents

\cleardoublepage
% \phantomsection
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\cleardoublepage
% \phantomsection
\addcontentsline{toc}{chapter}{\listtablename} 
\listoftables
\cleardoublepage

\addcontentsline{toc}{chapter}{\nameref{publication_statement}}
\chapter*{\label{publication_statement}Publications and statement of authorship}


The research reported is my own original work which was carried out in collaboration with
others as follows:

\paragraph{Chapter \ref{cha:intro}:} Written by Till Seuring. 

\paragraph{Chapter \ref{cha:review}:} Till Seuring was the lead author of a paper published as:
\\[12pt]
\noindent\fullcite{Seuring2015a}
\\[12pt]
\noindent Till Seuring, Marc Suhrcke and Olga Archangelidi designed the study. The search strategy was designed and executed by Till Seuring. Till Seuring and Olga Archangelidi screened the initial results and extracted the data from the primary studies. Till Seuring drafted the original manuscript which was critically reviewed by Olga Archangelidi and Marc Suhrcke.

\paragraph{Chapter \ref{cha:Mex1}:} Till Seuring was the lead author of a paper published as:
\\[12pt]
\noindent\fullcite{Seuring2015}
\\[12pt]
\noindent Till Seuring, Yevgeniy Goryakin and Marc Suhrcke  designed the study. Till Seuring analysed the data. Till Seuring drafted the original manuscript which was critically reviewed by Yevgeniy Goryakin and Marc Suhrcke.


\paragraph{Chapter \ref{cha:Mex2}:} Till Seuring was the lead author of a working paper published as: 
Seuring T., Serneels P., Suhrcke, M. (2016), ``The impact of diabetes on labour market outcomes in Mexico: a panel and biomarker data analysis.'' It has appeared as \textit{IZA Discussion Paper} 10123, and \textit{York University Centre for Health Economics Research Paper 134}, and has been submitted for publication.

Till Seuring, Pieter Serneels and Marc Suhrcke designed the study. Till Seuring analysed the data. Till Seuring drafted the original manuscript which was critically reviewed by Pieter Serneels and Marc Suhrcke.



\paragraph{Chapter \ref{cha:China}:} Till Seuring and Max Bachmann designed the study. Till Seuring analysed the data. Till Seuring drafted the original manuscript which was critically reviewed by Max Bachmann and Pieter Serneels.

\paragraph{Chapter \ref{cha:Discussion}:} Written by Till Seuring.
\cleardoublepage

\addcontentsline{toc}{chapter}{\nameref{acknowledgements}}
\chapter*{\label{acknowledgements}Acknowledgements}

Being able to write these lines today, at the end of my PhD, is the culmination of a lot of luck I have had in my life. I have been so lucky to be able to live a life of my own choosing that allows me to pursue my goals, to embark on new adventures and to be surrounded by those great people without whom this thesis would not have been possible. 

First of all, I am grateful for my the best supervisory team there is. I never would have thought that I would have the opportunity to work with such kind and bright people. Thank you Marc! Without you I may never would have pursued a PhD nor would I have found my way to UEA. I first met you at a talk you gave at the International Diabetes Congress, organized by my then employer. I asked you a couple of questions after your talk and we continued to chat for a few minutes afterwards and exchanged contact details. A few months later I contacted you regarding a PhD post at Cambridge, where you were involved in, and after some forth and back it became clear that there could be the opportunity to do a PhD with you at UEA. So it came that half a year later me and my wife actually moved to the fine city of Norwich. From then on it has been a great experience with a lot of work, but also a lot of fun. Thank you for always being responsive and helpful, interested in my work, providing great feedback and not hesitating one second to also support me from afar. The same has to be said of Pieter and Max, who became part of the supervisory team more or less at the midpoint of the PhD. Thank you Pieter for always pushing me to make my work better, for providing me with new ideas and for always being there when I had a question. I also enjoyed finding some common ground with you regarding the experience of having a wife from Latin America. Thank you Max for being a just as great primary supervisor as Marc had been before going to York, for being so enthusiastic, for introducing me to the world of epidemiology and for taking time out of your busy schedule to sit down with me for several hours to go through Stata code. Of course I also have to thank Olia and Yevgeniy for being such great co-authors on my first two papers. I also want to thank Marcello for helping me throughout the PhD.

Then there is Katey	 to who I am deeply indebted. I am so grateful that you provided me a home in Norwich and I thoroughly enjoyed going to the pub, or watching the possibly worst movie ever, with you and Gareth. I also want to thank Darren for being a good friend every time I came to Norwich. 

Being surrounded only by PhD students (in Norwich) or nobody who had a clue of what I was actually doing (back home), was not always easy. Therefore, I am happy to count on a number of friends that provided just the right mix of both worlds. In no particular order, I am especially grateful to Heiner for organizing the two-men conference in Kijkduin\footnote{See also \textcite{Schmittdiel2016}.} and for joining in on many similar events; to Christian for keeping me up to date on the latest political chatter in the German development sphere; to Alex for always keeping his doors in Moabit open for me; and to Frank for just being the lifelong friend everybody needs.

I cannot forget diabetes itself. You are one of the main reasons I am doing what I do. You not only send my blood sugar levels on a roller coaster but my life as well leaving me no choice but to learn to live with you. I can only be thankful to be born a part of the world where dealing with you is difficult but doable and I can live a ``normal'' life. Despite all the hard times you have given me, I am weirdly grateful that you handed me that personal invitation to pursue you in my research.  I hope that this thesis might be a small step towards a future where one day every person living with you will have the same possibilities I have had.

Of course this thesis would not have been possible without the best family in the world. Thank you Martin and Angelica for being the best \textit{suegros} and Annia for being the best \textit{cu\~{n}ada} I could have asked for. There is also my extended Mexican family, especially Wendy and Javier, which I am so happy to have. Being able to step into another world when visiting you is a blessing. It has given me much needed breaks from the life of a PhD student and being driven across Puebla to eat the best \textit{gorditas} with \textit{salsa verde con pipicha} is a very special treat. 

Anne and Elena, I am very fortunate to not only have such great sisters but to also be able to call you my friends. 
To my parents Martin and Petra: You are the reason that I am walking the face of this earth and have the chance to live this great life. Ich bin sehr glücklich Euch zu haben. Last, but certainly not least, a few words to my beautiful wife Ada and to my adorable little daughter Matilda, who both have been incredibly supportive even though they sometimes barely saw me during the final stages of this endeavour: Ich kann schwer beschreiben, wie sehr ich Euch liebe. Ihr seid das Beste in meinem Leben und mein größter Schatz. Las amo!

\cleardoublepage
\addcontentsline{toc}{chapter}{\nameref{abbreviations}}
\chapter*{\label{abbreviations}Abbreviations}

\input{acronyms}
\acresetall  %resets all acronyms



\chapter{\label{cha:intro}General introduction}
\include{Introduction}
\acresetall  %resets all acronyms	
\chapter{\label{cha:review}The economic costs of type 2 diabetes: a global systematic review}
\include{Chapter2}
\acresetall  %resets all acronyms
\chapter{\label{cha:Mex1}The impact of diabetes on employment in Mexico}
\section*{Pre-amble}

The systematic review in Chapter \ref{cha:review} identified a paucity of studies on the labour market impact of diabetes in developing countries. Further, even studies on \acp{HIC} did not provide much information regarding the heterogeneity of effects across different socioeconomic subgroups. There was no evidence on how diabetes may affect those in the formal compared to the informal labour market or across the wealth distribution. Further, it was unclear what the effects were for people unaware of their disease.

This study will use cross-sectional data from a large household survey in Mexico, assessing the impact of diabetes on employment probabilities. An \ac{IV} strategy inspired by preceding studies from \acp{HIC} is used to account for the potential endogeneity of diabetes due to unobserved heterogeneity. Especially personal characteristics such as ambition and family background could affect both the probability to develop diabetes, in particular type 2 diabetes, and the probability of being employed. The aim is to investigate if diabetes has a causal effect on employment probabilities and to provide evidence for the subgroup of those in the informal labour market and the relatively poor, populations of particular relevance in \acp{MIC}.
\include{Chapter3}
\acresetall  %resets all acronyms
\chapter{\label{cha:Mex2}The impact of diabetes on labour market outcomes in Mexico: a panel data and biomarker analysis}
\section*{Pre-amble}

This study builds on the results of the preceding chapter. Instead of using an \ac{IV} approach to address the issue of endogeneity, it takes advantage of the recently released third wave of the \ac{MxFLS} to allow the construction of a longitudinal data set containing three waves. This enables the use of panel data methods to arrive at a potentially causal interpretation of the estimates, without having to rely on an \ac{IV} approach.

Further, the study provides additional evidence for the effect of self-reported diabetes on wages and working hours in a developing country. Finally, it addresses another area identified by the systematic review in Chapter \ref{cha:review}. Using biomarker data it investigates in how diabetes effects the labour market outcomes of the large undiagnosed population, also providing information about whether findings based on self-reported diabetes can be used to infer on the entire population with diabetes. This should help to better interpret estimates using self-reported diabetes as provided in Chapter \ref{cha:Mex1}.

\include{Chapter4}
\acresetall  %resets all acronyms
\chapter{\label{cha:China} The relationship between diabetes, employment status and behavioural risk factors: An application of marginal structural models and fixed effects to Chinese panel data}
\section*{Pre-amble}

Chapters \ref{cha:Mex1} and \ref{cha:Mex2} provided evidence of the adverse impact of self-reported diabetes on employment probabilities in Mexico. However, if this is also the case in other \acp{MIC} is unclear. Chapter \ref{cha:China} adds to this using panel data covering a period of rapid economic transition in China, again estimating the relationship of diabetes and employment status. Moreover, it provides information about the ability of people with diabetes achieving changes in behavioural risk factors important for the prevention of diabetes complications, as studies have shown that smoking cessation and weight loss after a diagnosis can have beneficial effects on blood glucose control and the risk of complications. Importantly, it not only addresses time-invariant unobserved heterogeneity by using individual level fixed effects as in the previous chapter, but also accounts for the potential effects of time-variant confounding by using marginal structural models.

This method is widely applied in epidemiology and able to account for confounding over time, where prior outcomes can affect the current treatment, for example the previous employment status affects the current diabetes status. This potential source of bias has been assumed to not exist in previous studies, but could potentially have biased the estimate of the effect of diabetes on labour market outcomes. This chapter thereby makes several contributions compared to the previous chapters: It provides information about the robustness of the identified relationship of diabetes on employment status by using an alternative estimation strategy in a different setting, thereby also taking into account the potential effect of behavioural risk factors, and it gives first evidence in how far people with diabetes in China are able to change their behavioural risk factors after a diabetes diagnosis.


%}
\acresetall  %resets all acronyms
\include{Chapter5}
%\comment{
\acresetall  %resets all acronyms
\chapter{\label{cha:Discussion}Discussion and conclusions} 
\acresetall  %resets all acronyms
\include{Conclusion}
\printbibliography[heading=bibintoc]
\chapter*{\label{cha:Appendix}Appendices}
\addcontentsline{toc}{chapter}{\nameref{cha:Appendix}}
\renewcommand{\thechapter}{{\Roman{chapter}}} 
\renewcommand{\thetable}{A\arabic{table}} %changes chapter number to roman
\renewcommand{\thefigure}{A\arabic{figure}} %changes chapter number to roman
\setcounter{chapter}{0}  % sets chapter counter to zero to start numbering again for appendices with roman numbering
\setcounter{table}{0}
\setcounter{figure}{0}
\include{Appendix}
%}
\cleardoublepage
\end{document}          
